Finally Tagless, Partially Evaluated

### 2008-02-14 tagless/tagless-final.tex
typechecking experiments
I took care of all simple comments.

@@

We attack the problem of tagless (staged) typed-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

\section{A tagless compiler (or, a staged interpreter)}\label{compiler}
Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$
as values of type |('c, |$t$|) code|, where |'c| is the
environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run time.
Bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of three
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

We switch to the data type
\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code3}
module P1 : Symantics = struct
  let int (x:int) = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code3}
However, we stumble on functions.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

@@

Any partial evaluator must decide how much to unfold recursion\ifshort.
Our code na\"\i vely unfolds |fix|
\else\
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, unfold |fix| rather than residualize
\fi
whenever the argument is static.
\ifshort
In the accompanying source code is a conservative
alternative |P.fix| that unfolds recursion only once, then residualizes.
\else
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
\fi
Many sophisticated approaches have been developed to decide how much to unfold
\citep{jones-partial}, but this issue is orthogonal to our presentation.
\ifshort\else
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.
\fi

Given this implementation of~|P|, our running example
\ifshort
\texttt{let module E\,=\,EX(P) in E.test1 ()}
\else
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
\fi
evaluates to
\ifshort
|{P.st = Some true; P.dy = .<true>.}|
\else
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
\fi
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\ifshort
\texttt{let module E = EX(P) in E.testpowfix7}
\else
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
\fi
gives the desired result
\ifshort\vspace*{-0.7em}\fi
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
\end{code}
\ifshort\vspace*{-0.7em}\fi

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the compiler\fi~|C|).  To
distinguish between static and dynamic terms, one might define
\ifshort
\texttt{data P0~t~= S0~(R~t)~}\Verb+|+\texttt{ E0~(C~t)}.
\else
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
\fi
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
\ifshort\else
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add     e1      e2  = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Thus, the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

\fi
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |C Bool|), the second a static term (of type |R a|) and the third a
dynamic term (of type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.\ifshort\else
(By the way, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
f = app (lift f)~}.)\fi

@@

We switch to the Haskell data type
\ifshort \BUseVerbatim{P1} \else \UseVerbatim{P1} \fi
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
\ifshort
|abstr1 (P1 _ dyn) = dyn| of type
|abstr1 :: P1 t -> C t|
\else
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
\fi
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\ifshort
\vspace*{-0.6em}
\begin{code3}
instance Symantics P1 where
  int x = P1 (Just (int x)) (int x)
  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code3}
\vspace*{-0.6em}
\else
\begin{code}
instance Symantics P1 where
  int  x = P1 (Just (int  x)) (int  x)
  bool b = P1 (Just (bool b)) (bool b)

@@

  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing
                    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}
\fi
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\ifshort\vspace*{-0.6em}\fi
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code}
\ifshort\vspace*{-0.6em}\fi
unlike applying |P1| to |Int| or |Bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.
\ifshort\else The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.\fi

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\citep{oliveira-typecase}.  These
methods are equivalent, and here we use GADTs; |incope1.hs|
in the accompanying source code shows the latter.
\ifshort
\vspace*{-0.7em}
\begin{code3}
data P t where
  VI :: Int  -> P Int;                VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
\end{code3}
\vspace*{-0.7em}
\else
We introduce a GADT with four data constructors.
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code}
\fi
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  \ifshort\else However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|.  \fi We define a function
\ifshort |abstr|, \fi like |abstr1| above\ifshort,\fi\
to extract a future-stage computation from a value of type |P t|.
\ifshort\else
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

>>>>

%% TODO: add a ref 
%%  Lars Birkedal's master thesis
%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
%% and thank Eijiro Sumii

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{S:compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

\section{A tagless compiler (or, a staged interpreter)}\label{S:compiler}

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type~$t$
as values of type |('c, |$t$|) code|, where |'c| is the
environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
within a bracket and specifies that the expression~$e$ must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
the future-stage code value~$e$ by compiling and linking it at run time.
Bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of three
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

We switch to the data type
\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always available, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code3}
module P1 : Symantics = struct
  let int (x:int) = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code3}
However, we stumble on functions.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, it unfolds |fix| rather than residualize it.
\ifshort
In the accompanying source code is a conservative
alternative |P.fix| that unfolds recursion only once, then residualizes.
\else
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
\fi
Many sophisticated approaches have been developed to decide how much to unfold
\citep{jones-partial}, but this issue is orthogonal to our presentation.
\ifshort\else
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.
\fi

Given this implementation of~|P|, our running example
\ifshort
\texttt{let module E\,=\,EX(P) in E.test1 ()}
\else
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
\fi
evaluates to
\ifshort
|{P.st = Some true; P.dy = .<true>.}|
\else
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
\fi
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
\ifshort
\texttt{let module E = EX(P) in E.testpowfix7}
\else
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
\fi
gives the desired result
\ifshort\vspace*{-0.7em}\fi
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
\end{code}
\ifshort\vspace*{-0.7em}\fi

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add     e1      e2  = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Thus, the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |C Bool|), the second a static term (of type |R a|) and the third a
dynamic term (of type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}. (By the way, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
f = app (lift f)~}.)

@@

We switch to the Haskell data type
\UseVerbatim{P1}
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always available, we
can now define the polymorphic function
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code}
instance Symantics P1 where
  int  x = P1 (Just (int  x)) (int  x)
  bool b = P1 (Just (bool b)) (bool b)

@@

  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing
                    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code}
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code}
unlike applying |P1| to |Int| or |Bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type. The connection with CPS
is not an accident, as we shall see in~\S\ref{S:CPS}.

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\citep{oliveira-typecase}.  These
methods are equivalent, and here we use GADTs; |incope1.hs|
in the accompanying source code shows the latter.
We introduce a GADT with four data constructors.
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|.  We define a function
like |abstr1| above to extract a future-stage computation from a 
value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

We may now make |P| an instance of
|Symantics| and implement the partial evaluator as follows. We elide
|mul|, |leq|, |if_|, and |fix|.
\begin{code}
instance Symantics P where
  int x               = VI x
  bool b              = VB b
  add (VI n1) (VI n2) = VI (n1 + n2)
  add e1 e2           = E (add (abstr e1) (abstr e2))
  lam                 = VF
  app (VF f) ea       = f ea
  app (E f)  ea       = E (app f (abstr ea))
\end{code}
The implementations of |int|, |bool|, and |add| are like 
in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
which just wraps the HOAS function |f|. 
We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
|app ef ea| checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app| above: the type |P t| has four constructors, of which the pattern in
|app| matches only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed, the metalanguage implementation could reason thus:
if we use inductive families (as in Coq) or logical
  frameworks with canonical forms (as in Twelf with its coverage checker),
  we can prove the pattern matching to be exhaustive.
Then again, the metalanguage implementation may not reason thus:
GHC cannot and issues warnings.
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The
goal is to make this exhaustiveness \emph{syntactically} apparent.
